Automated iterative forecasting for the Portal Project

### 2018-02-01 manuscript/portal_iterative_forecasting_ms.Rmd
Edits from Hao

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of our fundamental understanding of ecology [@clark2001; @tallis2006; @diaz2015; @dietze2017]. Since Clark et al. [-clark2001] called for a more central role of forecasting in ecology, an increasing number of ecological forecasts are being published. However, most of these forecasts are made once, published, and never assessed or updated. Without assessment, we have limited information on how much confidence to place in our predictions; without regular updates, forecasts lack the most up-to-date information as conditions change [@dietze2018]. This lack of both regular assessment and active updating has limited the progress of ecological forecasting and hindered our ability to make useful and reliable predictions. For ecological forecasting to mature as a field, we need to change how we produce and interact with forecasts, creating a more dynamic interplay between model development, prediction generation, and incorporation of new data and information [@dietze2018].

@@

![Figure 1. Stages of the forecasting pipeline. To go from raw data to forecast presentation involves a number of stages, each of which required unique tasks, tools and infrastructure. While each stage has unique tasks associated with it, the stages of our pipeline are also interdependent. Our pipeline is primarily a linear structure where outputs from one stage form the inputs for the subsequent stage. All stages require code written in R [@R2017] ](figs/pipeline_fig.png).

@@

A core component of iterative near-term forecasting is the regular rerunning of the forecasting pipeline. This can be conducted manually---with a human making sure all code is run and all tables and files are updated---or automatically by having the computer conduct those tasks. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the tasks needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple software developers working on the same software. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Because of the widespread use of CI in software development, several CI services already exist. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time); alternative services that can be used to run code on local or cloud-based computational infrastructure are available, such as Drone (http://try.drone.io/) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 2).

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when a package one of our models relies on, `tscount` [@liboschik2015], was temporarily removed from CRAN (the R package repository) and, therefore, could not be installed in the usual way. This broke our forecasting pipeline because we could no longer run models that used that package. To minimize issues with changes in software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything you need to run some piece of software, including the operating system. Once created, a software container is basically a time capsule, it contains all the software dependencies in the exact state used to develop and run the software. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in your container. We use an existing platform, Docker [@Merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to update to new package versions after testing and making any necessary changes to the data processing and analysis code. We use a container created by the Rocker project which is a Docker image with many important R packages (i.e. tidyverse) pre-installed  [@boettiger2017]. We use add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

New data is double-entered into Microsoft Excel using the "data validation" feature. The two versions are then compared in an R script to control for errors in data entry. Quality control (QC) checks written in R using the `testthat` R package [@wickham2011] are run on the data to test for validity and consistency both within the new data and between the new and archived data. The local use of the QC scripts to flag problematic data greatly reduces the time spent error-checking and ensures that the quality of data is consistent. The data is then submitted to a GitHub-based data repository. GitHub (https://github.com/) is a software development tool for managing computer code development, but we have also found it useful for data management. Changes to data can be tracked through version control, and additions and changes to the data can be monitored through pull requests (notifications that someone has a modified version of the data to contribute). QA/QC checks are automatically rerun on the submitted data using continuous integration to ensure that these checks have been run and that no avoidable errors reach the official version of the dataset.

@@

The Portal Project has a long history of making its data publicly available, which means that anyone can use it for forecasting or other projects. Historically the publication of the data was conducted through data papers [@ernest2009,@ernest2016], the most common approach in ecology; however, this approach caused years of data latency. Recently, the project has switched to posting data directly to a public GitHub repository with a CC0 license (Figure 1). This immediate posting reduces that data latency to less than one month and, therefore, makes meaningful iterative near-term forecasting possible for not only our group but other interested parties, as well.

@@

Ideally, iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). New models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available (Figure 2). We use a plugin infrastructure to allow new models to be easily added to the system. Each model is a script that takes in standardized inputs and returns standardized outputs (Figure 3). All models are run by the main forecasting code at each update, and the standardized outputs are combined to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure allows flexibility in all aspects of the modeling process, making it easier to explore new modeling approaches, and allows new models that fit the data well to immediately improve the ensemble forecast.

![Figure 3. Demonstration of plugin infrastructure where each model script (represented here as AR1.R) uses data provided by the core forecasting code (represented here as rodent.csv) and returns its forecasts in a predefined structure that is consistent across models (represented here as 2017_12_08forecasts.csv) to allow cross-model evaluations.](figs/plugin_fig.gif)

To allow flexibility in what the models predict, we designed a forecast output format that allows us to store relatively generic forecasts. Each forecast output file contains the date being forecast, the collection date of the data used for fitting the models, the date the forecast was made, the state variable being forecast (e.g., rodent biomass, the abundance of a species), and the forecast value and associated uncertainty of that forecast (Figure 3). Our goal is to store a variety of different forecasts in a common format with the hope of this being a useful starting point for developing a standard for storing ecological forecasts more generally.

Forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics in the future. In addition to evaluating the actual forecasts, we also use hindcasting (forecasting on already collected data) to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12 month period can immediately be used to evaluate the model. Hindcasting can be conducted using data that was already been collected, thus allowing model comparison of large numbers of hindcasts and giving insight into which models make the best forecasts without needing to wait for new data to be collected. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate.

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2017; @harris2018] (Figure 1). This archiving serves as a form of pre-registration for model predictions, helping facilitate unbiased interpretation of model performance. We explored three major repositories for archiving our forecasts: FigShare (https://figshare.com/), Zenodo (https://zenodo.org/), and Open Science Framework (https://osf.io/). While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs (i.e., interfaces that allow automated ways interacting with these websites) and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to the GitHub repository from the CI server. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. There is an existing one-time process for linking our forecasting repository on  GitHub with Zenodo. Once this link is created, each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" of our repository. This triggers the Zenodo-GitHub integration, which automatically archives the resulting forecasts and the code that generated them under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). Through this process, we automatically archive every forecast made with a documented history of the archive. While this approach is functional because everything in the repository is archived when a new forecast is made, these archives have complicated folder structures, making it more harder for users to find and access the forecasting results. Finding simpler approaches to automated archiving would increase the transparency of the forecase results and their ease of use by external researchers.

@@

While building this system was facilitated by the use of existing technological solutions, there were still a number of challenges in making existing tools work for automated iterative forecasting. Continuous integration is designed primarily for running automated tests on software, not for running a coordinated forecasting pipeline. As a result, extra effort was sometimes necessary to figure out how to get these systems to work properly in non-standard situations, like running code that was not part of a software package. In addition, hosted continuous integration solutions, like Travis, provide only limited computational resources. As the number and complexity of the models we fit has grown, we have had to continually invest effort in reducing our total compute time so we can stay within these limits. Finally, we found no satisfactory existing solution for archiving our results. All approaches we tried had limitations when it came to automatically generating publicly versioned archives of forecasts on a repeated basis. Overall, we found existing technology to be sufficient to the task, but it required greater expertise and a greater investment of time than is ideal. Tool development to reduce the effort required for scientists to set up their own short-term forecasting systems would clearly be useful, but our efforts show that it is possible for scientists to use existing tools to develop initial iterative systems as a method for both advancing scientific understanding and developing proof of concept forecasting systems. By developing these systems so that they are producing automated iterative forecasts, it will be easier to convert these systems into fully operationalized forecasting systems that are relied on for decision making [@dietze2018, other paper by operationalization co-authors].

>>>>

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of our basic understanding of how ecology operates [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-clark2001] called for a more central role of forecasting in ecology, and since then an increasing number of ecological forecasts are being published. However, most of these forecasts are made once, published, and never assessed or updated. Without assessment, we have limited information on how much confidence to place in our predictions; without regular updates, forecasts lack the most up-to-date information as conditions change [@dietze2018]. This lack of both regular assessment and active updating has limited the progress of ecological forecasting and hindered our ability to make useful and reliable predictions. More regular updating and assessment will advance ecological forecasting as a field by speeding up the identification of the best models for forecasting specific issues in specific ecosystems and contribute overall to our understanding of how to best design forecasting approaches for ecology. For ecological forecasting to mature as a field, we need to change how we produce and interact with forecasts, creating a more dynamic interplay between model development, prediction generation, and incorporation of new data and information [@dietze2018].

@@

![Figure 1. Stages of the forecasting pipeline. To go from raw data to forecast presentation involves a number of stages, each of which requires unique tasks, tools and infrastructure. While each stage has unique tasks associated with it, the stages of our pipeline are also interdependent. Our pipeline is primarily a linear structure where outputs from one stage form the inputs for the subsequent stage. All stages require code written in R. ](figs/pipeline_fig.png).

@@

A core aspect of iterative near-term forecasting is the regular rerunning of the forecasting pipeline. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the computer scripts needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple developers working on the same code base. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time). Because of the widespread use of CI in software development, alternative services that can run code on local or cloud-based computational infrastructure also exist (e.g., Drone (http://try.drone.io/)) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 2).

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when a package one of our models relies on, `tscount` [@liboschik2015], was temporarily removed from CRAN (the R package repository) and, therefore, could not be installed in the usual way. This broke our forecasting pipeline because we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything you need to run some piece of software, including the operating system. Once created, a software container is basically a time capsule, it contains all the software dependencies in the exact state used to develop and run the software. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in your container. We use an existing platform, Docker [@Merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to update to new package versions after testing and making any necessary changes to the data processing and analysis code. We use a container created by the Rocker project which is a Docker image with many important R packages (i.e. tidyverse) pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

New data is double-entered into Microsoft Excel using the "data validation" feature. The two versions are then compared in an R script to control for errors in data entry. Quality control (QC) checks using the `testthat` R package [@wickham2011] are run on the data to test for validity and consistency both within the new data and between the new and archived data. The local use of the QC scripts to flag problematic data greatly reduces the time spent error-checking and ensures that the quality of data is consistent. The data is then uploaded to the GitHub-based Portal Data repository. GitHub (https://github.com/) is a software development tool for managing computer code development, but we have also found it useful for data management. On GitHub, changes to data can be tracked through the Git version control system which logs all changes made to any files in the repository. All updates to data are processed through "pull requests", which are notifications that someone has a modified version of the data to contribute. QA/QC checks are automatically rerun on the submitted data using continuous integration to ensure that these checks have been run and that no avoidable errors reach the official version of the dataset.

@@

The Portal Project has a long history of making its data publicly available, which means that anyone can use it for forecasting or other projects. Historically the publication of the data was conducted through data papers [@ernest2009,@ernest2016], the most common approach in ecology; however, this approach caused years of data latency. Recently, the project has switched to posting data directly to a public GitHub repository with a CC0 (i.e. no rights reserved) license (Figure 1). This immediate posting reduces that data latency to less than one month and, therefore, makes meaningful iterative near-term forecasting possible for not only our group but other interested parties, as well.

@@

Iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). Additionally, any new models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available (Figure 2). We use a plugin infrastructure to allow new models to be easily added to the system. Our approach is to treat each model as interchangable black boxes - we feed them all the same inputs and we expect them to generate the same structure for model outputs (Figure 3). Each iteration of the forecasting cycle, all models are run and the standardized outputs are combined into a single file to store the results of the different models' forecasts. This structure allows weighted ensemble models to be produced based on how well individual models fit the training data. Our hope is that by requiring all models to take the same input (though models may differ in how they use that input) and provide the same outputs, we've gained the capacity to add and compare very different types of models that we may not currently be able to foresee and design around. Our infrastrcture should allow easy addition of more complicated models and more sophisticated approaches to creating ensembles even though to date our models and ensemble-creating approaches are extremely basic.

![Figure 3. Demonstration of plugin infrastructure. All model scripts (represented here by the example AR1.R) are housed in a single folder. Each model script uses data provided by the core forecasting code (represented here by rodent.csv) and returns its forecast outputs in a predefined structure that is consistent across models (represented here by the example 2017_12_08forecasts.csv). Outputs from all models run on a particular date are combined into the same file (i.e. 2017_12_08forecasts.csv) to allow cross-model evaluations. Model outpus files are housed in a folder containing all forecast outputs from all previous dates to facilitate our arhciving process and and forecast assessment activities ](figs/plugin_fig.gif)

We designed our forecast output format to to store relatively generic forecasts information to allow flexibility in what the models predict (i.e. species abundances, biomass). Each forecast output file contains the date being forecast, the collection date of the data used for fitting the models, the date the forecast was made, the state variable being forecast (e.g., rodent biomass, the abundance of a species), and the forecast value and associated uncertainty of that forecast (Figure 3). Our goal is to store a variety of different forecasts in a common format with the hope of this being a useful starting point for developing a standard for storing ecological forecasts more generally.

Our forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics in the future. In addition to evaluating the actual forecasts, we also use hindcasting (forecasting on already collected data) to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12 month period can immediately be used to evaluate the model. Hindcasting can be conducted using data that was already been collected, thus allowing model comparison of large numbers of hindcasts and giving insight into which models make the best forecasts without needing to wait for new data to be collected. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate.

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2017; @harris2018] (Figure 1). Archiving serves as a form of pre-registration for model predictions, helping facilitate unbiased interpretation of model performance. We wanted our archives to not only be accessible to others but to also be a permanent record whose continued existence was not dependent on any particular researcher or funding stream. Storing forecast outputs in our forecasting GitHub repository was not sufficient for archival purposes because repositories can be deleted.  We explored three major repositories for archiving our forecasts: FigShare (https://figshare.com/), Zenodo (https://zenodo.org/), and Open Science Framework (https://osf.io/). While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs (i.e., interfaces that allow automated ways interacting with these websites) and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to the GitHub repository from the CI server. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. There is an existing one-time process for linking our forecasting repository on  GitHub with Zenodo. Once this link is created, each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" of our repository. This triggers the Zenodo-GitHub integration, which automatically archives the resulting forecasts and the code that generated them under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). Through this process, we automatically archive every forecast made with a documented history of the archive. While this approach is functional because everything in the repository is archived when a new forecast is made, these archives have complicated folder structures, making it more harder for users to find and access the forecasting results. Finding simpler approaches to automated archiving would increase the transparency of the forecase results and their ease of use by external researchers.

@@

While building this system was facilitated by the use of existing technological solutions, there were still a number of challenges in making existing tools work for automated iterative forecasting. Continuous integration is designed primarily for running automated tests on software, not for running a coordinated forecasting pipeline. As a result, extra effort was sometimes necessary to figure out how to get these systems to work properly in non-standard situations, like running code that was not part of a software package. In addition, hosted continuous integration solutions, like Travis, provide only limited computational resources. As the number and complexity of the models we fit has grown, we have had to continually invest effort in reducing our total compute time so we can stay within these limits. Finally, we found no satisfactory existing solution for archiving our results. All approaches we tried had limitations when it came to automatically generating publicly versioned archives of forecasts on a repeated basis. Overall, we found existing technology to be sufficient to the task, but it required greater expertise and a greater investment of time than is ideal. Tool development to reduce the effort required for scientists to set up their own short-term forecasting systems would clearly be useful, but our efforts show that it is possible for scientists to use existing tools to develop initial iterative systems as a method for both advancing scientific understanding and developing proof of concept forecasting systems. Forecasting systems such as ours can begin the process of making iterative forecasts while we wait for more integrated and easier to deploy platforms for iterative forecasting to be developed. Once more integrated platforms exist, it will be easier to convert current systems into fully operationalized forecasting systems that are relied on for decision making [@dietze2018, other paper by operationalization co-authors].
